---
title: "Project Report"
output: html_document
authors: Ashley Anderson, Cameron Stocker, Hexiang Zhang, & Tyler Kruzan
---

### Team Members:

Project Coordinator: Ashley Anderson

Team Members: Cameron Stocker, Hexiang Zhang, & Tyler Kruzan


### Dataset Overview

The dataset can be found on Kaggle or through this link:   
https://www.kaggle.com/PromptCloudHQ/imdb-data

```{r}

# Load dataset
movie = read.csv("IMDB-Movie-Data.csv")

```

### Background

Our choice of a data set came from IMDb and includes data about the 1,000 most popular movies that came out between 2006 and 2016. Useful variables include Title, Director, Year, Runtime, Rating, Metascore, & Revenue. There are some messy variables such as Description, Actors, & Genre that contain lengthy character strings that will need to be parsed and cleaned or simply omitted from the final data we use.

IMDb, which stands for Internet Movie Database, is an online source for data associated with films, TV shows, video games, and other streamable content. For each title, users can find information relating to the starring cast, producers, and plot summaries, as well as both critic and fan reviews. As of May 2019, IMDb had 83 million registered users and approximately 6 million titles, including 516,726 feature films. IMDb ratings are based on user ratings on a scale from 1 to 10. The raw ratings are then filtered and weighted, creating a weighted mean rating for each film. The formula for calculating the weighted rating is $W = \frac{R*v + C*m}{v+m}$, where $R$ is the raw average, $v$ is the number of votes, $m$ is the minimum votes required to be included in the Top 250 (which is currently 25,000), and $C$ is the mean vote for the reports (which is currently 7.0). Below is a table of all variables in the dataset as well as a description for each variable. 

| Dataset Variables  | Description      |
| ------------------ | ---------------- |
| Rank               | Movie Rank Order |
| Title              | Tite of the film |
| Genre              | Comma separated-list of genres used to classify the film |
| Description        | Brief one-sentence movie summary |
| Director           | The name of the film's director |
| Actors             | A comma-separated list of the main stars of the film |
| Year               | The year that the film released as an integer. |
| Runtime (Minutes)  | The duration of the film in minutes.|
| Rating             | Average user rating for the movie on a 0-10 scale |
| Votes              | Total number of ratings given per movie |
| Revenue (Millions) | Movie revenue in millions |
| Metascore          | An aggregated average of critic scores. Values are between 0 and 100. Higher scores represent positive reviews. |

### Pipeline

We begin by describing our plan of attack for the project. Beginning with the raw data set, we start with some exploration in order to see what needed to be fix or removed. We then move on to wrangling and cleaning the data based on what we found during the exploration.  

For analysis, we will examine a variety of questions. First we will find which actors and directors are most commonly involved with successful movies. We define success by looking at a movie's rating, revenue, or metascore. We use the separate method in the 'Actors' column to move it into multiple columns. From there, we find the average movie rating and revenue on an actor and director specific basis.

We also dive into what factors are correlated with a high rating and revenue. Specifically, we look at how runtime effects these success variables and if they have any significance in predicting rating or revenue. This can be done through ploting the dataset. 

We also examine relationships between different factors in a movie over time such as the year it was made and movie performances like runtime, votes, rating, and revenue to see how the components of top movies have changed over the years. This can also be done through plotting the dataset. 

Lastly, we explore how different genres perform in term of success; including ratings, revenue, and metascore. We also compare these success variables to each other to see which genre is most "successful" and which one is least "successful". 


### Data Exploration, Cleaning and Wrangling 

We can see from the structure of the data set that each variable type seems to be correct, so no changes were needed there. However, we noticed that there were some problems.

```{r message=FALSE, warning=FALSE}
#Install Packages
library(tidyverse)
library(ggplot2)
str(movie)
summary(movie)
head(movie)
```

First, we noticed a few NAs in the data set and decided to explore more. As we can see below, there are quite a few NAs in the Revenue column. Since this variable is important for our definition of success, we decided to remove all rows that contained any NAs. It was also crucial to do so in some of our analyses that used averages, as we did not want NAs to effect our results. 

```{r}
colSums(is.na(movie))
movie <- movie %>% drop_na()
```

Second, some of the variables are messy or need renamed. We decided to drop the Rank and Description columns because these variables are lengthy and difficult to use. Additionally, we do not believe they would add much value to our analysis. Next, we changed the variable names for Runtime and Revenue so that they did not include the units (minutes and millions) in the name. 

```{r}
movie = movie %>% select(-Rank, -Description)
names(movie)[c(6,9)] = c("Runtime", "Revenue")
```

Next, the genre variable shows multiple genres for each movie. In order for this variable to be useful, we need to separate them. We do this by separating the genres and then spreading the column out so each genre is a column. The movies recognized under a certain genre display a "1" value, and those that are not display a "0" value.  

```{r}
movie = separate_rows(movie, Genre, sep = ",")
movie = movie %>%
  mutate(yesno = 1) %>%
  distinct %>%
  spread(Genre, yesno, fill = 0)
```

Lastly, the Actor variable shows multiple actors for each movie. Again, for this column to be useful, we will need to isolate each actor for each movie. We do this by separating the "Actors" column into four different columns: Actor1, Actor2, Actor3, and Actor4.  

```{r}

movie = movie %>% separate("Actors", sep = ",", into = c("Actor1", "Actor2", "Actor3", "Actor4"))

```


### Data Analysis

#### Which actors and directors are most/least successful?

##### Actor Analysis

For our actor analysis, we used both the Revenue and Metascore variables to examine 'success'. 

```{r}
###Had to create four different datasets, one with each Actor variable
actor1 = movie %>% select(Actor1, Rating, Votes, Revenue, Metascore)
actor2 = movie %>% select(Actor2, Rating, Votes, Revenue, Metascore)
actor3 = movie %>% select(Actor3, Rating, Votes, Revenue, Metascore)
actor4 = movie %>% select(Actor4, Rating, Votes, Revenue, Metascore)
###Then renamed the Actor variable to be "Actor"
names(actor1)[c(1)] = c("Actor")
names(actor2)[c(1)] = c("Actor")
names(actor3)[c(1)] = c("Actor")
names(actor4)[c(1)] = c("Actor")
###Combined the 4 datasets through a row bind so that they could be grouped
actorSet = rbind(actor1, actor2, actor3, actor4)
actorSet = actorSet %>% filter(!is.na(Actor))
###Had to go through the actor variable and check for any empty spaces that were formed from the separate function.
for(i in 1:length(actorSet$Actor)){
  if(substr(actorSet$Actor[i], 1, 1) == " "){
    actorSet$Actor[i] = substr(actorSet$Actor[i], 2, nchar(actorSet$Actor[i]))
  }
}
###Group the actors together by average
actorGrouped = actorSet %>% group_by(Actor) %>% summarize(count = n(), Revenue = mean(Revenue),  Rating = mean(Rating), Votes = mean(Votes), Metascore = mean(Metascore))

###Decided to only examine actors with 5 or more movies to help prevent any outliers.
actorFinal = filter(actorGrouped, count>4)

actorRevenue = rbind(head(actorFinal[order(-actorFinal$Revenue),], 5), head(actorFinal[order(actorFinal$Revenue),], 5))

actorRevTrim = as.data.frame(actorRevenue)

actorRevTrim$Actor = factor(actorRevTrim$Actor, levels = actorRevTrim[order(actorRevTrim[,"Revenue"]),1])

ggplot(actorRevTrim) + geom_bar(aes(x = Actor, y = Revenue, fill = Revenue), stat = "identity") + coord_flip() + geom_label(aes(x=Actor,y = Revenue, label = count)) + labs(title = "Average Revenue by Actor", y = "Revenue (millions $)")
```

In the plot above, the numbers at the end of the bar represent the number of movies that actor was in. Josh Hutcherson had the highest revenue per movie, which is comes from being in the Hunger Games movies. The lowest, Charlotte Gainsbourg, was involved in three movies that didn't have a rating which makes it difficult for those movies to have much success in theatres.

```{r}
###Create the metascore data set
actorMetascore = rbind(head(actorFinal[order(-actorFinal$Metascore),], 5), head(actorFinal[order(actorFinal$Metascore),], 5))

actorMetTrim = as.data.frame(actorMetascore)

actorMetTrim$Actor = factor(actorMetTrim$Actor, levels = actorMetTrim[order(actorMetTrim[,"Metascore"]),1])
ggplot(actorMetTrim) + geom_bar(aes(x = Actor, y = Metascore, fill = Metascore), stat = "identity") + coord_flip() + geom_text(aes(x=Actor,y = Metascore, label = paste(count, "movies")), color = "white", size = 5, position = position_stack(vjust = 0.5)) + labs(title = "Average Metascore by Actor")
```

As you can see from the plot above Rooney Mara had the highest average metascore and Adam Sandler had the lowest. Although Adam Sandler is widely considered to be successful, he has been involved in some comedy movies that had very low ratings and this plot highlights that.

##### Director Analysis

For the director analysis, we used the Rating statistic to define 'success'.

```{r}
directorGrouped = movie %>% group_by(Director) %>% summarize(count = n(), Revenue = mean(Revenue),  Rating = mean(Rating), Votes = mean(Votes), Runtime = mean(Runtime), Metascore = mean(Metascore))

directorFinal = filter(directorGrouped, count>2)
directorRating = rbind(head(directorFinal[order(-directorFinal$Rating),], 5), head(directorFinal[order(directorFinal$Rating),], 5))
directorRatTrim = as.data.frame(directorRating)
directorRatTrim$Director = factor(directorRatTrim$Director, levels = directorRatTrim[order(directorRatTrim[,"Rating"]),1])

ggplot(directorRatTrim) + geom_bar(aes(x = Director, y = Rating, fill = Rating), stat = "identity") + coord_flip() + geom_text(aes(x=Director,y = Rating, label = paste(count, "movies")), color = "white", size = 5, position = position_stack(vjust = 0.5)) + labs(title = "Average Rating by Director")
```

As you can see from the plot above, the top 3 directors for average movie rating were Christopher Nolan, Quentin Tarantino, and Martin Scorsese. That is interesting because if you ask any who they think are some of the top directors in the world, you would likely hear these names come up, and this plot confirms that belief.

#### How is runtime correlated with high ratings and revenue?
```{r}

movie2 = movie %>%
    group_by(Runtime) %>%
    summarize(mean_rev = mean(Revenue))
movie2$group = ifelse(movie2$Runtime <= 30, 1,ifelse(movie2$Runtime <= 60, 2, ifelse(movie2$Runtime <= 90, 3, ifelse(movie2$Runtime <= 120, 4, ifelse(movie2$Runtime <= 150, 5, ifelse(movie2$Runtime <= 180, 6, 7))))))

ggplot(movie2, aes(x = Runtime, y = mean_rev, color=group)) + geom_point() + labs(title = "Average Revenue of Movies for each Runtime") + xlab("Movie Runtime (min)") + ylab("Revenue (mil)") + stat_smooth(method="lm", se = FALSE)

fit <- lm(mean_rev ~ Runtime, data = movie2)

cat("Adj R2 =",signif(summary(fit)$adj.r.squared, 5))
cat("y =",signif(fit$coef[[2]], 5), "x +",signif(fit$coef[[1]],5 ))
cat("P =",signif(summary(fit)$coef[2,4], 5))

movie2 = movie %>%
    group_by(Runtime) %>%
    summarize(mean_rat = mean(Rating))

movie2$group = ifelse(movie2$Runtime <= 30, 1,ifelse(movie2$Runtime <= 60, 2, ifelse(movie2$Runtime <= 90, 3, ifelse(movie2$Runtime <= 120, 4, ifelse(movie2$Runtime <= 150, 5, ifelse(movie2$Runtime <= 180, 6, 7))))))

ggplot(movie2, aes(x = Runtime, y = mean_rat, colour=group)) + geom_point() + labs(title = "Average Rating of Movies for each Runtime") + xlab("Movie Runtime (min)") + ylab("Rating") + stat_smooth(method="lm", se = FALSE)

fit <- lm(mean_rat ~ Runtime, data = movie2)

cat("Adj R2 =",signif(summary(fit)$adj.r.squared, 5))
cat("y =",signif(fit$coef[[2]], 5), "x +",signif(fit$coef[[1]],5 ))
cat("P =",signif(summary(fit)$coef[2,4], 5))

```

#### How do the number of movies created change over time?
```{r}
movie %>% 
  ggplot(aes(x = Year)) + geom_histogram(fill = "#e69b22") + labs(y = "Number of Movies", title = "Quantity of Movies Over the Years")
```

#### How do movie ratings compare over time?
```{r}
movie %>%
  ggplot(aes(y = Rating)) + geom_boxplot(fill = "#7afffb") +facet_grid(~Year) + labs(title = "Comparison of Ratings Over the Years")+ theme(axis.text.x = element_blank(), axis.ticks.x=element_blank())
```

#### How have movie revenues changed over the years?
```{r}
movie %>%
  ggplot(aes(x = Year, y = Revenue)) + geom_col(fill = "#42f54b") + labs(title = "Movie Revenue through the Years")
```

#### What genres are most successful?

In the graph below, there are three bar plots that represent the average revenue in millions by genre, the average metascore by genre, and the average rating by genre. This was done by creating a new data frame, which includes variables genre, average revenue, average rating, and average metascore. Each unique genre was represented once in the genre column, and for each unique genre, the averages for revenue, ratings, and metascore could be calculated using a for loop.

```{r fig.height=5, fig.width=10}

genres <- colnames(movie[,13:32])
movie2<- movie %>%
  select(genres,"Revenue","Metascore","Rating") 
movie3 <- data.frame("Genre"=genres, "Avg_Revenue"=NA, "Avg_Metascore"=NA, "Avg_Rating"=NA)
for(i in 1:20){
  total1 <- 0
  total2 <- 0
  total3 <- 0
  n <- 0
  for(j in 1:nrow(movie2)){
    if(movie2[j,i]==1){
      total1 <- total1 + movie2$Revenue[j]
      total2 <- total2 + movie2$Metascore[j]
      total3 <- total3 + movie2$Rating[j]
      n <- n + 1
    }
  }
  avg1 <- total1/n
  movie3$Avg_Revenue[i]=avg1
  avg2 <- total2/n
  movie3$Avg_Metascore[i]=avg2
  avg3 <- total3/n
  movie3$Avg_Rating[i]=avg3
}

g1 <- movie3 %>%
  arrange(desc(Avg_Revenue)) %>%
  ggplot(aes(x=reorder(Genre, Avg_Revenue),y=Avg_Revenue, fill=Genre)) + geom_bar(stat='identity') +
  coord_flip() + theme(legend.position = "none") + geom_text(aes(label=round(Avg_Revenue, digits=1)), position=position_dodge(width=0.9), hjust=1) +labs(title="Average Revenue by Genre", x="Genre", y="Average Revenue (millions)")

g2 <- movie3 %>%
  arrange(desc(Avg_Metascore)) %>%
  ggplot(aes(x=reorder(Genre, Avg_Metascore),y=Avg_Metascore, fill=Genre)) + geom_bar(stat='identity') +
  coord_flip() + theme(legend.position = "none") + 
  geom_text(aes(label=round(Avg_Metascore, digits=1)), position=position_dodge(width=0.9), hjust=1) +
  labs(title="Average Metascore by Genre", x="Genre", y="Average Metascore")


g3 <- movie3 %>%
  arrange(desc(Avg_Rating)) %>%
  ggplot(aes(x=reorder(Genre, Avg_Rating),y=Avg_Rating, fill=Genre)) + geom_bar(stat='identity')+
  coord_flip() + theme(legend.position = "none") + 
  geom_text(aes(label=round(Avg_Rating, digits=1)), position=position_dodge(width=0.9), hjust=1) +
  labs(title="Average Ratings by Genre", x="Genre", y="Average Ratings")

#install.packages("grid")
library(grid)
grid.newpage()

pushViewport(viewport(layout = grid.layout(nrow = 1, ncol = 3)))

define_region <- function(row, col){
  viewport(layout.pos.row = row, layout.pos.col = col)
} 

print(g1, vp = define_region(row = 1, col = 1)) 
print(g2, vp = define_region(row = 1, col = 2))
print(g3, vp = define_region(row = 1, col = 3))

```

Comparing the three barplots, it is obvious that the genre 'Animation' performed very well. It had the top revenue of any revenue, exceeding the second highest average revenue genre by about \$35 million for a total average of \$192.4 million. Not only did 'Animation' films do well in producing revenue, we can see that they were second in average metascores with an average score of 70.6 out of 100, and fourth in average ratings with a rating of 7.3 out of 10. This means 'Animation' films were generally enjoyed by both critics and IMDb fans.

Conversely, the 'Horror' genre did not perform as well. Comparing the barplots, we see that 'Horror' movies accumulated the smallest amount of revenue out of any genre, which was about \$37.4 million. Additionally, it performed among the bottom five average metascores, with an average score of 56.5 out of 100. Lastly, 'Horror' movies have the least favorable rating by IMDb fans, with an average rating of 6.3 out of 10. Overall, 'Horror' movies generally perform unsuccessfully. 

There are some interesting trends to point out in these graphs. Genres like 'Fantasy', 'Action', and 'Family' recieved success in revenue, as they are among the top six genres for average revenue. However, we can see that these three genres were among the bottom four in both average metascore and average ratings. Thus, generally speaking, 'Fantasy', 'Action', and 'Family' movies generate a lot of revenue, but are not well recieved. This could be because these are genres a majority of the population watch, including children, meaning it can earn a large revenue. But just because a large proportion of the population watches a movie from this genre does not necessarily mean the movie will be well recieved.

Another trend to point out is opposite of the trend above genres. Genres like 'History', 'War' and 'Biography' do not seem to accumulate a ton of revenue. However, they are among the top four for average metascores and they are the top three genres for average ratings. Unlike 'Fantasy', 'Action', and 'Family', generally 'History', 'War' and 'Biography' movies are very well recieved, but do not generate a lot of money. This could be because these three genres a not generally as popular as a lot of people may find them boring and unexciting, meaning they do not generate as much revenue.  

### Conclusion

One major take away from this project is that you cannot look at a single number when defining success. For one, when looking at which actors are most successful, there was no overlap in the top five actors who recieved the highest average revenue and the top five actors who recieved the highest average metascore. While some actors had success by making a lot of revenue from roles, others had success in their movies being well recieved by critics and fans. Similarly, when looking at which genres are most successful, there is sometimes a big difference in ranking between revenue, metascore, and ratings. For example, could see this in the fact that genres like 'Fantasy', 'Action', and 'Family' recieved high success in revenue but were among the bottom four in both average metascore and average ratings.

Through our analysis, we were able to discover why there are many ways to define the success of a movie. Going forward, it might be interesting to look into why certain actors, directors, and genres have higher success with revenue while others have higher success with critic and fan ratings. Additionally, it would be interesting to figure out why there is an increase in total number of movies being produced every year. 

### Team Contribution

* Ashley: In this project, we decided to separate the data analysis section and we all focused on different questions of interest we came up together as a team. I specifically analyzed how genres of a movie contribute to it's success, which was the last question analyzed above. I also organized our report document so we could be sure to follow the rubric and guidelines, as well as help to set up our presentation. 

* Tyler: 

* Hexiang: I worked on cleaning some of the data. I worked on organizing the genre section of our project. I used separate and one hot encoded the movies and their genres. I worked on looking at how runtime is related to high revenue and high ratings from users on IMDb. I worked on create our powerpoint presenation presented in class. 

* Cameron: I worked on cleaning some of the data. I separated the Actor variable into 4 different actor variables so that we could look at them individually. For analysis, I examined the which actors/directors were the most/least successful by using different variables. I believe the plots were very intuitive and effectively displayed the results we were looking for. I also was there to provide teammates with suggestions on how they could try to improve their analysis.